---
title: "midterm3_code"
author: "Gavrila Ang"
date: "4/26/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

Descriptive variables: neutrophils_abs, lymphs_abs, monocytes_abs, eos_abs, baso_abs, immature_grans_abs, CRP_MSD,SAA_MSD, sICAM.1_MSD, sVCAM.1_MSD
Observation: total_cholesterol

Clinical identifiers have been factorised for PCA analyses
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(psych)
library(ggfortify)
library(factoextra)
```

```{r Loading dataset}
# Import datasetset
dataset <- read.csv("hapichi_dataset.csv", header = T) 

# Filtering the datasetset to contain only the baseline visit, thereby removing intervention effects
# Selecting only the columns containing the variables I'm interested in, for easier PCA analyses
dataset <- dataset %>% filter(visit=="1") %>%  
  select(c("cholesterol_total",	
           "neutrophils_abs", "lymphs_abs",	"monocytes_abs",	"eos_abs",	"baso_abs",
           "immature_grans_abs",
           "CRP_MSD",	"SAA_MSD", "sICAM.1_MSD",	"sVCAM.1_MSD"
           )) %>%
  na.omit(T)# Omit any participant that did not have complete clinical records

# Creating a tchol_bin column, that categorises participants according to their total cholesterol levels
# These categories are based on LabCorp reference intervals for their total cholesterol assays
# LabCorp link: https://www.labcorp.com/tests/001065/cholesterol-total
# Upper limit of 400 was used as an arbitary cutoff value 

dataset$tchol_bin <- cut(dataset$cholesterol_total, breaks=c(0, 200, 240, 400), labels = c("Acceptable", "Borderline", "High"), right = FALSE)

# Save these cateogical variables assigned to each participant into a new column
dataset$tchol_bin <- as.factor(dataset$tchol_bin)

# Relocate tchol_bin column as the second column in the datasetframe
dataset <- relocate(dataset, tchol_bin, .after=cholesterol_total)

# View how many participants are in each interval group
# Unbalanced groups, would like to have more people in the "High" category
# Exclusionary criteria of the study most likely lead to few participants with "High" total cholesterol levels
dataset %>% group_by(tchol_bin) %>% tally()
```

```{r Preliminary dataset analyses}
# Produce a pairs plot with the descriptive variables ONLY
pairs.panels(dataset[,c(-1:-2)])

pairs.panels(dataset[,c("neutrophils_abs", "monocytes_abs","sICAM.1_MSD", "sVCAM.1_MSD", "CRP_MSD", "SAA_MSD")])

# ICAM and VCAM are correlated (0.77)
# CRP and SAA are somewhat correlated (0.58)
# CRP and ICAM are somewhat correlated (0.53)
# SAA and ICAM are somewehat correlated (0.41)
# Monocytes and neutrophils are somewhat correlated (0.49)

# The high collinearity between ICAM and VCAM would prevent the inclusion of both variables in a linear model
# However, PCA combines these variables to create uncorrelated principle components and removes multi-collinearity for potential downstream analyses
```


```{r Performing a PCA}
# Perform a PCA, using centered and scaled trait dataset
# dataset is centred and scaled so that no single descriptive variable will dominate the PCA procedure
dataset_pca <- prcomp(dataset[,-1:-2], center=TRUE, scale.= TRUE)

# Show the proportion of variance explained by each principal component
# A lot of principle components were generated and it took up to PC6 to explain 82% of the variation within the dataset!
summary(dataset_pca)

# Produce a screeplot to visualise the summary statistics above
screeplot(dataset_pca, main = '', las = 1, cex.lab = 1.5)
```


```{r Creating a biplot for the PCA}
# Produce a biplot using the ggfortify package autoplot() function, showing the participants coloured by total cholesterol category, and arrows showing the descriptive variables included in the PCA.

# Looks overwhelming; however, it should be noted that the same "type" of measurements have arrows with smaller angles.
# This indicates higher correlation between these descriptive variables and is in corroboration of the pairs plot results
# Make the text larger for the presentation slides
autoplot(dataset_pca, data = dataset, colour = 'tchol_bin', loadings=TRUE,
size = 3, loadings.label = TRUE, loadings.label.size=8) +
  xlim(-0.3, 0.45) + 
      theme(text = element_text(size = 30, face = "bold"),
        axis.title = element_text(size = 30, face = "bold"),
        axis.text = element_text(size = 30, face = "bold"))
```

```{r Performing factovizcontribute with the PCA}
# Extract the values for the variables' contributions to the principal components.
var = get_pca_var(dataset_pca)

# Show a screeplot of the contributions of each descriptive variable
# Only view contributions to PC1-PC6, as these explained ~83% of the variation in the datasetset.
# Make the text larger for the presentation slides
fviz_contrib(dataset_pca, choice='var', axes = 1:6) +
    theme(text = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold"),
        axis.text = element_text(size = 20, face = "bold"))

# Retain only descriptive variables with bars above the red line
# These are the variables that largely contribute to PC1-PC6
# In other words, these are the descriptive variables that explain the majority 
# of the variance found in the total_cholesterol values and must be kept for the
# remainder of the analyses
reduced.dataset = dataset[,c(1,2,8,6,11,12)]
```

```{r Performing a PCA with the reduced datasetset}
# Fit a PCA with the reduced datasetset, using centered and scaled trait dataset
reduced.dataset_pca <- prcomp(reduced.dataset[,-1:-2], center=TRUE, scale.= TRUE)

# Show the proportion of variance explained by each principal component
# By reducing the number of descriptive variables, we see that the PCA is better
# able to explain the variation in the observation with fewer principle components
# Only 4 PCAs are created, and it only takes 2 to explain 71% of the variation, yay!
summary(reduced.dataset_pca)

# Produce a screeplot to visualise the summary statistics above
screeplot(reduced.dataset_pca, main = '', las = 1, cex.lab = 1.5)
```

```{r Creating a biplot for the reduced PCA}
# Generate biplots with the new PCA, showing the participants coloured by total cholesterol category, and arrows showing the descriptive variables included in the PCA.
# Add ellipses around each clinical group

# Show arrows and labels of variables, points of individuals
# Make the text larger for the presentation slides
fviz_pca_biplot(reduced.dataset_pca, habillage=reduced.dataset$tchol_bin,
                addEllipses = TRUE, ellipse.level=0.95,
                label="var",
                labelsize = 12,
                pointsize = 4) +
  xlim(-7, 5) + 
  theme(text = element_text(size = 40, face = "bold"),
        axis.title = element_text(size = 40, face = "bold"),
        axis.text = element_text(size = 40, face = "bold"))

# Show only individuals only
# Make the text larger for the presentation slides
fviz_pca_biplot(reduced.dataset_pca, habillage=reduced.dataset$tchol_bin,
                addEllipses = TRUE, ellipse.level=0.95,
                geom.ind=c("point"),
                invisible="var")

# Show variables factor map of PCA
# Control variable colors using their contributions
# Make the text larger for the presentation slides
fviz_pca_var(reduced.dataset_pca, col.var="contrib",
             labelsize = 10) + 
  xlim(-1.5, 1.0) + ylim (-1.0, 1.0) + 
  scale_color_gradient2(low="blue", mid="black", high="red", midpoint = 25) + theme_bw() +
    theme(text = element_text(size = 30, face = "bold"),
        axis.title = element_text(size = 30, face = "bold"),
        axis.text = element_text(size = 30, face = "bold"))
```

Conclusion:
The inflammatory descriptive variables included in the PCA were not able to definitively group the clinical population in accordance to "total cholesterol" LabCorp reference levels. In other words, the clinical population, identified by their "total cholesterol" levels, do not appear to have distinct inflammatory profiles.
